\documentclass[12pt]{article}

% ── Packages ──────────────────────────────────────────────────────────────────
\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{times}
\usepackage{microtype}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}

% ── Document ──────────────────────────────────────────────────────────────────
\begin{document}

\begin{center}
  {\large \textbf{A Physics-Informed Fourier Neural Operator for\\[4pt]
  Rapid Emulation of 1D Dynamic Earthquake Rupture}}\\[10pt]

  Aimran\,$^1$\\[4pt]
  {\small $^1$Department of Mathematical Sciences,
           New Mexico State University, Las Cruces, NM}\\[4pt]
  {\small \textit{NMSU--UTEP Joint Mathematics Workshop, April 2026}}
\end{center}

\vspace{6pt}
\hrule
\vspace{10pt}

\noindent
\textbf{Abstract.}\;
Numerical simulation of earthquake rupture dynamics is computationally
expensive, requiring the repeated solution of coupled elastic wave and
friction equations over large parameter spaces.
This makes tasks such as uncertainty quantification, parameter inversion,
and real-time hazard assessment prohibitively costly with classical solvers.
We present a proof-of-concept \emph{Physics-Informed Neural Operator}
(PINO) framework that learns to emulate a 1D dynamic rupture simulator
orders of magnitude faster than the underlying physics solver.

The forward simulator discretizes two elastic half-spaces separated by a
fault interface governed by \emph{Slip-Weakening} (SW) friction.
The governing velocity-stress system,
\begin{equation}
  \frac{\partial v}{\partial t} = \frac{1}{\rho}\frac{\partial s}{\partial x},
  \qquad
  \frac{\partial s}{\partial t} = \mu\,\frac{\partial v}{\partial x},
\end{equation}
is solved with fourth-order Summation-by-Parts (SBP) finite differences and
Runge-Kutta time integration.
The fault traction obeys the slip-weakening law
$\tau = \sigma_n\bigl[\alpha_s - (\alpha_s-\alpha_d)\min(D/D_c,1)\bigr]$,
enforced weakly via Simultaneous-Approximation-Term (SAT) penalties.
A Numba-JIT implementation achieves a \textbf{1\,868$\times$} speedup over
pure-Python serial execution (0.086\,s per simulation on a single CPU thread).

The neural operator maps a four-dimensional physical parameter vector
$\mathbf{p} = (\tau_0,\,\alpha_s,\,\alpha_d,\,D_c)$ — background stress,
static and dynamic friction coefficients, and critical slip distance — to the
full space-time solution
$\mathcal{G}_\theta(\mathbf{p}) = \bigl(v_l(x,t),\,s_l(x,t),\,v_r(x,t),\,s_r(x,t)\bigr)$
on a $64\times 64$ spatial-temporal grid.
The architecture is a \textbf{2D Fourier Neural Operator} (FNO-2D):
parameter channels are broadcast over the grid and lifted into a
$d_v\!=\!64$ latent space; four spectral-convolution blocks filter the
leading 16 Fourier modes in each dimension; and a two-layer projection
head maps to the four output fields.
The model has 8.4\,million trainable parameters.

A training dataset of 300 simulations (200 train / 50 validation / 50 test)
is generated via Latin Hypercube Sampling (LHS) of the parameter space with
physical validity constraints (nucleation condition $\tau_0 > \alpha_s\sigma_n$,
velocity-weakening condition $\alpha_d < \alpha_s$).
The model is trained with an \emph{AdamW} optimizer and cosine learning-rate
schedule using a relative $L^2$ data loss,
\begin{equation}
  \mathcal{L}_{\text{data}}
  = \frac{1}{N}\sum_{i=1}^{N}
    \frac{\|\mathcal{G}_\theta(\mathbf{p}_i) - u_i\|_2}{\|u_i\|_2}.
\end{equation}
Preliminary results show rapid convergence: the validation loss drops from
$\sim\!1.0$ at initialization to $\mathbf{0.023}$ by epoch~50
($\sim$2.3\% relative field error), with training continuing toward
200 epochs.
The PINO framework further supports a \emph{physics residual} loss
evaluated exactly in Fourier space — exploiting the identity
$\partial_x u \leftrightarrow ik_x\hat{u}$ — enabling PDE-constrained
training without finite-difference approximation of derivatives.

This work demonstrates that Fourier Neural Operators can accurately capture
the nonlinear wave-friction dynamics of dynamic rupture from a small
supervised dataset, offering a pathway toward GPU-accelerated surrogate
models for earthquake source inversion and stochastic ground-motion
simulation.

\vspace{10pt}
\noindent
\textbf{Keywords:} Fourier Neural Operator, Physics-Informed Neural Operator,
dynamic rupture, earthquake simulation, Slip-Weakening friction,
surrogate modeling, SBP-SAT.

\end{document}
